---
title: "R in Data Science - Breakout Session - Campus DevCon Summit"
author: "Jose Marie Antonio Minoza"
date: '2023-09-08'
output: html_document
---

# Introduction to R

## What is R?

R is a powerful and versatile programming language and environment for statistical computing and data analysis. It was developed by statisticians and data scientists for a wide range of data-related tasks, from data manipulation and visualization to advanced statistical modeling.

## Why use R?

R is widely used in academia, industry, and research for several reasons:

-   Open-source: R is free and open-source, making it accessible to anyone.
-   Active community: A large and active community of users and developers provides support and contributes packages.
-   Extensive libraries: R has a rich ecosystem of packages for various data analysis and visualization tasks.
-   Data manipulation: R excels at data manipulation and transformation.
-   Statistical analysis: R offers a wide range of statistical techniques and models.
-   Data visualization: R has excellent tools for creating informative data visualizations.

# Data Types in R

## Numeric

Numeric data types are used to represent numbers, both integers and real numbers (floating-point).

```{r}
# Numeric variables 
x <- 5 
y <- 3.14
```

## Character

Character data types are used to store text and strings.

```{r}
# Character variables
name <- "John"
message <- "Hello, world!"
```

## Logical

Logical data types represent binary values - either TRUE or FALSE.

```{r}
# Logical variables
is_raining <- TRUE
is_sunny <- FALSE
```

## Factors

Factors are used to represent categorical data with predefined levels.

```{r}
# Creating a factor
gender <- factor(c("Male", "Female", "Male", "Female"))
```

## Data Frames

Data frames are used to store tabular data with rows and columns.

```{r}
# Creating a data frame
data <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 22)
)
```

# Matrices in R

## Creating Matrices

Matrices are two-dimensional arrays in R.

```{r}
# Creating a matrix
matrix_data <- matrix(1:6, nrow = 2, ncol = 3)
```

## Matrix Operations

You can perform various operations on matrices, including addition, multiplication, and transposition.

```{r}

# Matrix operations
mat_a <- matrix(1:6, nrow = 2, ncol = 3)
mat_b <- matrix(7:12, nrow = 2, ncol = 3)
mat_sum <- mat_a + mat_b
mat_prod <- mat_a %*% t(mat_b)
```

## Indexing and Subsetting

You can access elements and subsets of matrices using indexing.

```{r}

# Matrix indexing and subsetting
first_row <- matrix_data[1, ]
second_col <- matrix_data[, 2]
subset_matrix <- matrix_data[1:2, 2:3]
```

# Functions in R

## Built-in Functions

R provides numerous built-in functions for various purposes.

```{r}
# Built-in functions
mean_value <- mean(c(1, 2, 3, 4, 5))
sqrt_value <- sqrt(16)
```

## Creating User-defined Functions

You can define your own functions in R using the function() keyword.

```{r}
# User-defined function
my_function <- function(x, y) {
  result <- x^2 + y^2
  return(result)
}
```

## Function Arguments

Functions can have arguments that allow you to pass values to them.

```{r}
# Function with arguments
calculate_area <- function(length, width) {
  area <- length * width
  return(area)
}
```

# The Apply Family of Functions

The apply family of functions in R is used for applying a function to the rows or columns of a matrix or data frame. It includes functions like apply, lapply, sapply, and mapply.

## apply()

The apply() function applies a specified function to either rows or columns of a matrix.

```{r}
# Example of apply() to calculate row sums
row_sums <- apply(matrix_data, 1, sum)
```

## lapply()

The lapply() function applies a function to each element of a list or vector and returns a list.

```{r}
# Example of lapply() to square each element of a list
numbers <- list(a = 1, b = 2, c = 3)
squared_numbers <- lapply(numbers, function(x) x^2)
```

## sapply()

The sapply() function is similar to lapply() but simplifies the result to a vector or matrix if possible.

```{r}
# Example of sapply() to square each element of a list
numbers <- list(a = 1, b = 2, c = 3)
squared_numbers <- sapply(numbers, function(x) x^2)
```

## mapply()

The mapply() function is used for multiple input vectors and applies a function element-wise.

```{r}
# Example of mapply() to calculate the sum of two vectors element-wise
vector1 <- c(1, 2, 3)
vector2 <- c(4, 5, 6)
sum_vector <- mapply(function(x, y) x + y, vector1, vector2)
```

# Machine Learning in R



# R Shiny

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

### Widget

```{r, echo=FALSE}
# shinyAppDir(
#   system.file("examples/06_tabsets", package = "shiny"),
#   options = list(
#     width = "100%", height = 550
#   )
# )
```

### Tabs

```{r, echo=FALSE}
# shinyAppDir(
#   system.file("examples/06_tabsets", package = "shiny"),
#   options = list(
#     width = "100%", height = 550
#   )
# )
```
