---
title: "R in Data Science - Breakout Session - Campus DevCon Summit"
author: "Jose Marie Antonio Minoza"
date: '2023-09-08'
output: html_document
---

# Introduction to R

## What is R?

R is a powerful and versatile programming language and environment for statistical computing and data analysis. It was developed by statisticians and data scientists for a wide range of data-related tasks, from data manipulation and visualization to advanced statistical modeling.

## Why use R?

R is widely used in academia, industry, and research for several reasons:

-   Open-source: R is free and open-source, making it accessible to anyone.
-   Active community: A large and active community of users and developers provides support and contributes packages.
-   Extensive libraries: R has a rich ecosystem of packages for various data analysis and visualization tasks.
-   Data manipulation: R excels at data manipulation and transformation.
-   Statistical analysis: R offers a wide range of statistical techniques and models.
-   Data visualization: R has excellent tools for creating informative data visualizations.

# Data Types in R

## Numeric

Numeric data types are used to represent numbers, both integers and real numbers (floating-point).

```{r}
# Numeric variables 
x <- 5 
y <- 3.14
```

## Character

Character data types are used to store text and strings.

```{r}
# Character variables
name <- "John"
message <- "Hello, world!"
```

## Logical

Logical data types represent binary values - either TRUE or FALSE.

```{r}
# Logical variables
is_raining <- TRUE
is_sunny <- FALSE
```

## Factors

Factors are used to represent categorical data with predefined levels.

```{r}
# Creating a factor
gender <- factor(c("Male", "Female", "Male", "Female"))
```

# Operators

```{r}
a <- 1 + 1 # addition
b <- 2 * 9 # multiplication
c <- 3 - 1 # subtraction
d <- a / c # division
e <- b %% 5 # modulus
```

# Special Data Types
```{r}
NA
NaN
NULL
```
```{r}
1/0
90000^200
```
```{r}
Inf * -2
Inf - Inf
```
```{r}
-2023/Inf
2023/Inf
-2023/-Inf
2023/-Inf
```


```{r}
-2023/0
2023/0
```
## Data Frames

Data frames are used to store tabular data with rows and columns.

```{r}
# Creating a data frame
data <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 22)
)
```

# Matrices in R

## Creating Matrices

Matrices are two-dimensional arrays in R.

```{r}
# Creating a matrix
matrix_data <- matrix(1:6, nrow = 2, ncol = 3)
```

## Matrix Operations

You can perform various operations on matrices, including addition, multiplication, and transposition.

```{r}

# Matrix operations
mat_a <- matrix(1:6, nrow = 2, ncol = 3)
mat_b <- matrix(7:12, nrow = 2, ncol = 3)
mat_sum <- mat_a + mat_b
mat_prod <- mat_a %*% t(mat_b)
```

## Indexing and Subsetting

You can access elements and subsets of matrices using indexing.

```{r}

# Matrix indexing and subsetting
first_row <- matrix_data[1, ]
second_col <- matrix_data[, 2]
subset_matrix <- matrix_data[1:2, 2:3]
```


```{r}
A <- matrix(data = c(1,2,3,4,5,6,7,8),nrow = 2, ncol = 4)
A
```


```{r}
A[1,2] # A[row,column]
A[,2] # return all rows of Column 2
A[1,] # return all columns of Row 1
```
```{r}
C <-  matrix(data = 1:100,nrow = 5, ncol = 20)
C
```

```{r}
C[1,1:20][C[1,1:20]<10]
C[,1][C[,1]>10]
```


# Functions in R

## Built-in Functions

R provides numerous built-in functions for various purposes.

```{r}
# Built-in functions
mean_value <- mean(c(1, 2, 3, 4, 5))
sqrt_value <- sqrt(16)
```


```{r}
# Logarithms 
log(x=144,base=12)
log(x=216,base=6)

# Exponentials
exp(x=3)
log(x=20.17)
```

```{r}
x <- c(2,0,2,3) # Creates vector with elements 2,0,0,4
y <- c(1,6,2,9)
x + y # Sums elements of two vectors
```
```{r}
x * 4 # Multiplies elements
```
```{r}
sqrt(x) # Function applies to each element
```
```{r}
vector <- c(1,2,3,4,5)
length(vector)
```


## Creating User-defined Functions

You can define your own functions in R using the function() keyword.

```{r}
# User-defined function
my_function <- function(x, y) {
  result <- x^2 + y^2
  return(result)
}
```

## Function Arguments

Functions can have arguments that allow you to pass values to them.

```{r}
# Function with arguments
calculate_area <- function(length, width) {
  area <- length * width
  return(area)
}
```


```{r}
calculate_area(4, 4)
```
```{r}

l <- seq(1,4)
w <- rep(5,4)
l
w
```
```{r}
calculate_area(l, w)
```


# The Apply Family of Functions

The apply family of functions in R is used for applying a function to the rows or columns of a matrix or data frame. It includes functions like apply, lapply, sapply, and mapply.

## apply()

The apply() function applies a specified function to either rows or columns of a matrix.

```{r}
# Example of apply() to calculate row sums
row_sums <- apply(matrix_data, 1, sum)
```

## lapply()

The lapply() function applies a function to each element of a list or vector and returns a list.

```{r}
# Example of lapply() to square each element of a list
numbers <- list(a = 1, b = 2, c = 3)
squared_numbers <- lapply(numbers, function(x) x^2)
```

## sapply()

The sapply() function is similar to lapply() but simplifies the result to a vector or matrix if possible.

```{r}
# Example of sapply() to square each element of a list
numbers <- list(a = 1, b = 2, c = 3)
squared_numbers <- sapply(numbers, function(x) x^2)
```

## mapply()

The mapply() function is used for multiple input vectors and applies a function element-wise.

```{r}
# Example of mapply() to calculate the sum of two vectors element-wise
vector1 <- c(1, 2, 3)
vector2 <- c(4, 5, 6)
sum_vector <- mapply(function(x, y) x + y, vector1, vector2)
```

# Data Science in R

## Visualization

### Basic Graphics (Base R)

Basic graphics in R allow you to create simple plots using functions like plot, hist, and boxplot. Let's create some basic visualizations.

#### Scatterplot

```{r}
head(iris, 5)
```


```{r}
# Scatterplot of Sepal Length vs. Sepal Width
plot(iris$Sepal.Length, 
     iris$Sepal.Width, 
     xlab = "Sepal Length", 
     ylab = "Sepal Width", 
     main = "Scatterplot of Sepal Length vs. Sepal Width")
```
```{r}
pacman::p_load('ggplot2')
# Scatterplot using ggplot2
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point() +
  labs(x = "Sepal Length", y = "Sepal Width", title = "Scatterplot of Sepal Length vs. Sepal Width")
```
```{r}
# Histogram of Petal Length
hist(iris$Petal.Length, xlab = "Petal Length", main = "Histogram of Petal Length")
```


```{r}
# Histogram of Petal Length using ggplot2
ggplot(iris, aes(x = Petal.Length)) +
  geom_histogram(binwidth = 0.2, fill = "blue", color = "black") +
  labs(x = "Petal Length", title = "Histogram of Petal Length")
```
```{r}
# Boxplot of Sepal Length by Species
boxplot(Sepal.Length ~ Species, 
        data = iris, 
        xlab = "Species", 
        ylab = "Sepal Length", 
        main = "Boxplot of Sepal Length by Species")

```
```{r}
# Boxplot of Sepal Length by Species using ggplot2
ggplot(iris, aes(x = Species, y = Sepal.Length)) +
  geom_boxplot(fill = "green", color = "black") +
  labs(x = "Species", y = "Sepal Length", title = "Boxplot of Sepal Length by Species")

```
## Regression Analysis

In this tutorial, we will explore regression analysis in R. Regression is a statistical technique used to model the relationship between a dependent variable and one or more independent variables. We will cover the following topics:

1. Introduction to Regression
2. Loading and Preparing the Data
3. Simple Linear Regression
4. Multiple Linear Regression
5. Model Evaluation
6. Conclusion

### 1. Introduction to Regression

Regression analysis helps us understand how the value of a dependent variable changes when one or more independent variables change. It is commonly used for prediction and understanding relationships between variables.

### 2. Loading and Preparing the Data

#### Loading Data

Let's start by loading a dataset. For this tutorial, we'll use a sample dataset like the `mtcars` dataset, which contains information about car attributes.

```{r}
# Load the mtcars dataset
data(mtcars)
```

```{r}
head(mtcars)
```

###3. Simple Linear Regression

Simple linear regression is used when there is a linear relationship between the dependent variable and a single independent variable. Let's perform simple linear regression.

```{r}

# Fit a simple linear regression model
model_simple <- lm(mpg ~ hp, data = mtcars)

# Print the regression summary
summary(model_simple)
```
### 4. Multiple Linear Regression

Multiple linear regression extends simple linear regression to include multiple independent variables. Let's perform multiple linear regression.

```{r}
# Fit a multiple linear regression model
model_multiple <- lm(mpg ~ hp + wt + qsec, data = mtcars)

# Print the regression summary
summary(model_multiple)
```
### 5. Model Evaluation

To evaluate the regression models, we can use various metrics such as R-squared, adjusted R-squared, and residuals analysis.

```{r}

# Evaluate the simple linear regression model
rsq_simple <- summary(model_simple)$r.squared
adj_rsq_simple <- summary(model_simple)$adj.r.squared

# Evaluate the multiple linear regression model
rsq_multiple <- summary(model_multiple)$r.squared
adj_rsq_multiple <- summary(model_multiple)$adj.r.squared
```


```{r}
rsq_simple
adj_rsq_simple
```
```{r}
rsq_multiple
adj_rsq_multiple
```
```{r}
saveRDS(model_multiple, "final_regression_model.RDS")
```


# R Shiny

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

### Widget

```{r, echo=FALSE}
# shinyAppDir(
#   system.file("examples/06_tabsets", package = "shiny"),
#   options = list(
#     width = "100%", height = 550
#   )
# )
```

### Tabs

```{r, echo=FALSE}
# shinyAppDir(
#   system.file("examples/06_tabsets", package = "shiny"),
#   options = list(
#     width = "100%", height = 550
#   )
# )
```
